---
title: "Dashboard"
output: 
  flexdashboard::flex_dashboard:
    orientation: rows
---

```{r echo = FALSE}
#{r options} applies to chunks of code 

knitr::opts_chunk$set(echo = FALSE) #sets global options to apply for every chunk in the file
library(flexdashboard)

```
Importing the Data {data-navmenu="Task 2"}
===================================== 

Row {data-height=100}
-------------------------------------
### Importing the Data
```{r echo = TRUE, eval=FALSE}
data <- read.csv("C:/Users/Malcolm/Downloads/SeoulBikeData.csv", header=FALSE)
```

Row
-------------------------------------
### Cleaning and Processing the data
``` {r echo = TRUE, eval=FALSE}
# remove 1st row
data <- data[-c(1), ]
# rename columns
colnames(data) <- c("date","RentCount", "Hour", "Temp", "Hum", "Wind", 'Vis', "Dew","SolRad","Rain", "Snow","Season", "Holiday","Function")
data <- type.convert(data, as.is= TRUE, na.strings = "NA")

data$Function[data$Function == 'No'] <- FALSE
data$Function[data$Function == 'Yes'] <- TRUE

data$Holiday[data$Holiday == 'No Holiday'] <- FALSE
data$Holiday[data$Holiday == 'Holiday'] <- TRUE

data$date <- dmy(data$date)
#settings the correct type
data$RentCount <- as.integer(as.character(data$RentCount))
data$Hour <- as.numeric(as.character(data$Hour))
data$Temp <- as.numeric(as.character(data$Temp))
data$Hum <- as.numeric(as.character(data$Hum))
data$Vis <- as.numeric(as.character(data$Vis))
data$Dew <- as.numeric(as.character(data$Dew))
data$Wind <- as.numeric(as.character(data$Wind))
data$SolRad <- as.numeric(as.character(data$SolRad))
data$Rain <- as.numeric(as.character(data$Rain))
data$Snow <- as.numeric(as.character(data$Snow))
```

Data Imputation {data-navmenu="Task 2"}
===================================== 
### The data provided contained null values as can be seen below:

```{r}
# finds the null values
library(lubridate)
library(knitr)

# import csv
data <- read.csv("C:/Users/Malcolm/Downloads/SeoulBikeData.csv", header=FALSE)
# remove 1st row
data <- data[-c(1), ]
# rename columns
colnames(data) <- c("date","RentCount", "Hour", "Temp", "Hum", "Wind", 'Vis', "Dew","SolRad","Rain", "Snow","Season", "Holiday","Function")
data <- type.convert(data, as.is= TRUE, na.strings = "NA")

data$Function[data$Function == 'No'] <- FALSE
data$Function[data$Function == 'Yes'] <- TRUE

data$Holiday[data$Holiday == 'No Holiday'] <- FALSE
data$Holiday[data$Holiday == 'Holiday'] <- TRUE

data$date <- dmy(data$date)
data$RentCount <- as.integer(as.character(data$RentCount))
data$Hour <- as.numeric(as.character(data$Hour))
data$Temp <- as.numeric(as.character(data$Temp))
data$Hum <- as.numeric(as.character(data$Hum))
data$Vis <- as.numeric(as.character(data$Vis))
data$Dew <- as.numeric(as.character(data$Dew))
data$Wind <- as.numeric(as.character(data$Wind))
data$SolRad <- as.numeric(as.character(data$SolRad))
data$Rain <- as.numeric(as.character(data$Rain))
data$Snow <- as.numeric(as.character(data$Snow))

# finds the null values
null_data <- data[!complete.cases(data),]
kable(null_data, caption = "Null values")
```
As a data imputation strategy we opted to use the '(Rounded) Mean or Moving Average or Median Value' strategy. This implementation was done by getting the average value of the missing data of the same Season and Hour and replacing the missing value with it. After the data imputation was complete, the null values were replaced:

```{r}
# finds the null values
library(utf8)
library(pillar)
library(lubridate)
library(dplyr)
library(knitr)

# import csv
data <- read.csv("C:/Users/Malcolm/Downloads/SeoulBikeData.csv", header=FALSE)
# remove 1st row
data <- data[-c(1), ]
# rename columns
colnames(data) <- c("date","RentCount", "Hour", "Temp", "Hum", "Wind", 'Vis', "Dew","SolRad","Rain", "Snow","Season", "Holiday","Function")
data <- type.convert(data, as.is= TRUE, na.strings = "NA")

data$Function[data$Function == 'No'] <- FALSE
data$Function[data$Function == 'Yes'] <- TRUE

data$Holiday[data$Holiday == 'No Holiday'] <- FALSE
data$Holiday[data$Holiday == 'Holiday'] <- TRUE

data$date <- dmy(data$date)
data$RentCount <- as.integer(as.character(data$RentCount))
data$Hour <- as.numeric(as.character(data$Hour))
data$Temp <- as.numeric(as.character(data$Temp))
data$Hum <- as.numeric(as.character(data$Hum))
data$Vis <- as.numeric(as.character(data$Vis))
data$Dew <- as.numeric(as.character(data$Dew))
data$Wind <- as.numeric(as.character(data$Wind))
data$SolRad <- as.numeric(as.character(data$SolRad))
data$Rain <- as.numeric(as.character(data$Rain))
data$Snow <- as.numeric(as.character(data$Snow))

# finds the null values
null_data <- data[!complete.cases(data),]

for(i in 1:nrow(null_data)) 
{
  
  for(j in 1:ncol(null_data))
  {
    # if there is a null in the column we replace it with the seasons hourly average
    if(is.na(null_data[i,j]))
    {
      # gets column name of missing value
      col_name = (colnames(null_data)[j])
      #gets the hour where the missing value is
      hour = null_data[i, "Hour"]
      #gets all data where the season and hour match that of the missing value
      season_hour_data = filter(data, Season == null_data[i, "Season"] & Hour == hour) 
      #gets average value
      hourly_average = mean(season_hour_data[,col_name], na.rm = TRUE)
      if(col_name == "RentCount")
      {
        hourly_average = floor(hourly_average)
      }
      #replaces the missing value with the average value
      null_data[rownames(null_data)[i],col_name] <- hourly_average
    }
  }
}

kable(null_data, caption = "Imputated Null values")
```

The code for the imputation is as follows:
```{r echo = TRUE, eval = FALSE}
for(i in 1:nrow(null_data)) 
{
  for(j in 1:ncol(null_data))
  {
    # if there is a null in the column we replace it with the seasons hourly average
    if(is.na(null_data[i,j]))
    {
      # gets column name of missing value
      col_name = (colnames(null_data)[j])
      #gets the hour where the missing value is
      hour = null_data[i, "Hour"]
      #gets all data where the season and hour match that of the missing value
      season_hour_data = filter(data, Season == null_data[i, "Season"] & Hour == hour) 
      #gets average value
      hourly_average = mean(season_hour_data[,col_name], na.rm = TRUE)

      if(col_name == "RentCount")
      {
        hourly_average = floor(hourly_average)
      }
      #replaces the missing value with the average value
      data[rownames(null_data)[i],col_name] <- hourly_average
    }
  }
}
```


Deviation {data-navmenu="Data Analysis"}
===================================== 

Row
-------------------------------------

### Graph

```{r}
library(utf8)
library(pillar)
library(lubridate)
library(dplyr)
library(ggplot2)
library(reshape2)
library(data.table)

# import csv
data <- read.csv("C:/Users/Malcolm/Downloads/SeoulBikeData.csv", header=FALSE)
# remove 1st row
data <- data[-c(1), ]
# rename columns
colnames(data) <- c("date","RentCount", "Hour", "Temp", "Hum", "Wind", 'Vis', "Dew","SolRad","Rain", "Snow","Season", "Holiday","Function")
data <- type.convert(data, as.is= TRUE, na.strings = "NA")

data$Function[data$Function == 'No'] <- FALSE
data$Function[data$Function == 'Yes'] <- TRUE

data$Holiday[data$Holiday == 'No Holiday'] <- FALSE
data$Holiday[data$Holiday == 'Holiday'] <- TRUE

data$date <- dmy(data$date)
data$RentCount <- as.integer(as.character(data$RentCount))
data$Hour <- as.numeric(as.character(data$Hour))
data$Temp <- as.numeric(as.character(data$Temp))
data$Hum <- as.numeric(as.character(data$Hum))
data$Vis <- as.numeric(as.character(data$Vis))
data$Dew <- as.numeric(as.character(data$Dew))
data$Wind <- as.numeric(as.character(data$Wind))
data$SolRad <- as.numeric(as.character(data$SolRad))
data$Rain <- as.numeric(as.character(data$Rain))
data$Snow <- as.numeric(as.character(data$Snow))

# finds the null values
null_data <- data[!complete.cases(data),]
# list rows that contain the null values

# Task 2
for(i in 1:nrow(null_data)) 
{
  
  for(j in 1:ncol(null_data))
  {
    # if there is a null in the column we replace it with the seasons hourly average
    if(is.na(null_data[i,j]))
    {
      # gets column name of missing value
      col_name = (colnames(null_data)[j])
      #gets the hour where the missing value is
      hour = null_data[i, "Hour"]
      #gets all data where the season and hour match that of the missing value
      season_hour_data = filter(data, Season == null_data[i, "Season"] & Hour == hour) 
      #gets average value
      hourly_average = mean(season_hour_data[,col_name], na.rm = TRUE)

      if(col_name == "RentCount")
      {
        hourly_average = floor(hourly_average)
      }
      #replaces the missing value with the average value
      data[rownames(null_data)[i],col_name] <- hourly_average
    }
  }
}
```

```{r}
# Task3

# candle 3

# group data by month
monthly <- data %>%  group_by(month = lubridate::floor_date(date, "month")) %>%
  summarize(average = mean(RentCount), sd = sd(RentCount))

# convert from date to month name
monthly$month <- months(as.Date(monthly$month)) 
# order by month
monthly <- monthly %>% mutate(month = factor(month, levels=month.name)) %>% arrange(month)

# plot average and sd per month
p <- ggplot(monthly, aes(x=month, y=average)) + ylab("Rent Count") + xlab("Month") + ggtitle("Average and Standard Deviation of Rented Bike Count for each month") +
  geom_bar(stat="identity",fill="blue") +
  geom_errorbar(aes(ymin=average-sd, ymax=average+sd), width=.3) +
  geom_point(size=2)

p
```

Row
-------------------------------------

### Description
FILL HERE

Seasonal Statistics {data-navmenu="Data Analysis"}
===================================== 

Row
-------------------------------------
### Graph

```{r}

# gets list of the different Seasons in the dataset
seasons <- unique(data$Season)

# creates empty dataframe
seasonal_hourly = data.frame()

# loops through the seasons
for(i in seasons)
{
  
  # filters the data to show only of season i
  x <- filter(data, Season == i) %>% 
    # groups the data by the hour
    group_by(Hour) %>% 
    # gets the mean RentCount for each hour
    summarise(average = mean(RentCount))
  
  #transposition of x 
  y <- transpose(x)
  # adds the season that corresponds to the data
  y <- cbind(y, i)
  # names the columns
  colnames(y) <- c(0:23, "Season")
  # removes the first row which held the column names
  y <- y[-c(1), ]
  # adds the season row to the seasonal hourly
  seasonal_hourly = rbind(seasonal_hourly, y)

}
#moves season column to front (not necessary but easier to understand when viewing)
seasonal_hourly <- seasonal_hourly %>% relocate(Season)
# melts the seasonal_hourly dataframe so as to be able to plot
melted <- melt(seasonal_hourly, id.vars="Season")
# plots a line for each Season
p <- melted %>% ggplot(aes(x=variable,y=value, color = Season, group=Season)) + xlab("Hour") + ylab("Average Rent Count") + ggtitle("Rented Bike Count by Hour of the day across season") + 
  geom_line(linewidth=1.2)

p
```

Row
-------------------------------------
### Description

FILL HERE

Daily Statistics {data-navmenu="Data Analysis"}
===================================== 

Row
-------------------------------------

### Graph
```{r}
# selecting only required columns
weekday_data <- data %>% select(date, RentCount, Hour)
# convert date to day of week
weekday_data$date <- weekdays(weekday_data$date)
# remove saturday and sunday
weekday_data <- filter(weekday_data, date != "Saturday" & date != "Sunday")

# get the different days of week in the data
days_of_week <- unique(weekday_data$date)

# create a new empty dataframe
grouped_weekday_data = data.frame()

# loops through the different days
for(i in days_of_week)
{
  # filters data to show only of the day i
  x <- filter(weekday_data, date == i) %>%
    # groups data by the hour
    group_by(Hour) %>% 
    # gets mean for each hour
    summarise(average=mean(RentCount))
  
  # transposition of x
  y <-transpose(x)
  # adds the day that corresponds to the date
  y <- cbind(y,i)
  # renames columns
  colnames(y) <- c(0:23, "Day")
  # removes the first row which held the column names
  y <-y[-c(1),]
  # adds the day row to the grouped weekday data
  grouped_weekday_data = rbind(grouped_weekday_data, y)
}
#moves day column to front (not necessary but easier to understand when viewing)
grouped_weekday_data <- grouped_weekday_data %>% relocate(Day)
# melts the grouped_weekday_data so as to be able to plot
melted_grouped_weekday_data <- melt(grouped_weekday_data, id.vars="Day")

# plots line for each day of week
p <- melted_grouped_weekday_data %>% ggplot(aes(x=variable, y=value, color=Day, group=Day)) +
  xlab("Hour") + ylab("Average Rent Count") + ggtitle("Rented Bike Count by Hour of the day across weekdays") + 
  geom_line(linewidth=1.2)

p
```

Row
-------------------------------------

### Description
FILL HERE

Summer and Winter Comparisons {data-navmenu="Data Analysis"}
===================================== 


Row 
-------------------------------------

### Graph 1
```{r}
#filters and only keeps required columns
summer_winter_data <- data %>% filter(Season == "Summer" | Season =="Winter") %>% select(date, Hour, RentCount, Season)

# gets hourly averages
summer_winter_hourly_average <- summer_winter_data %>% group_by(Season) %>% summarise(average=mean(RentCount))

# gets totals for day
summer_winter_daily_average <- summer_winter_data %>% group_by(date) %>% summarise(Season=Season, total=sum(RentCount)) %>%
# gets daily averages
  group_by(Season) %>% summarise(average=mean(total))

# plots daily averages
p1 <- ggplot(summer_winter_daily_average, aes(x=Season, y=average)) + 
  ylab("Average Rent Count") + xlab("Season") + ggtitle("Daily Averages of Rented Bike Count for Summer and Winter") +
  geom_bar(position = "dodge", stat = "identity")

p1
```

### Graph 2
```{r}
# plots hourly averages
p2 <- ggplot(summer_winter_hourly_average, aes(x=Season, y=average)) + 
  ylab("Average Rent Count") + xlab("Season") + ggtitle("Hourly Averages of Rented Bike Count for Summer and Winter") +
  geom_bar(position = "dodge", stat = "identity")
p2
```

Row
-------------------------------------

### Description
FILL HERE