---
title: "Dashboard"
output: 
  flexdashboard::flex_dashboard:
    orientation: rows  
    vertical_layout: scroll
---

```{r echo=FALSE, eval=TRUE}
#{r options} applies to chunks of code 

knitr::opts_chunk$set(echo = FALSE) #sets global options to apply for every chunk in the file
library(flexdashboard)
library(utf8)
library(pillar)
library(lubridate)
library(dplyr)
library(ggplot2)
library(reshape2)
library(data.table)
library(knitr)

# import csv
data <- read.csv("C:/Users/Malcolm/Downloads/SeoulBikeData.csv", header=FALSE)
# remove 1st row
data <- data[-c(1), ]
# rename columns
colnames(data) <- c("date","RentCount", "Hour", "Temp", "Hum", "Wind", 'Vis', "Dew","SolRad","Rain", "Snow","Season", "Holiday","Function")
data <- type.convert(data, as.is= TRUE, na.strings = "NA")

data$Function[data$Function == 'No'] <- FALSE
data$Function[data$Function == 'Yes'] <- TRUE

data$Holiday[data$Holiday == 'No Holiday'] <- FALSE
data$Holiday[data$Holiday == 'Holiday'] <- TRUE

data$date <- dmy(data$date)
data$RentCount <- as.integer(as.character(data$RentCount))
data$Hour <- as.numeric(as.character(data$Hour))
data$Temp <- as.numeric(as.character(data$Temp))
data$Hum <- as.numeric(as.character(data$Hum))
data$Vis <- as.numeric(as.character(data$Vis))
data$Dew <- as.numeric(as.character(data$Dew))
data$Wind <- as.numeric(as.character(data$Wind))
data$SolRad <- as.numeric(as.character(data$SolRad))
data$Rain <- as.numeric(as.character(data$Rain))
data$Snow <- as.numeric(as.character(data$Snow))

# finds the null values
null_data <- data[!complete.cases(data),]
filled_null_data <- data[!complete.cases(data),]
# list rows that contain the null values

# Task 2
for(i in 1:nrow(null_data)) 
{
  
  for(j in 1:ncol(null_data))
  {
    # if there is a null in the column we replace it with the seasons hourly average
    if(is.na(null_data[i,j]))
    {
      # gets column name of missing value
      col_name = (colnames(null_data)[j])
      #gets the hour where the missing value is
      hour = null_data[i, "Hour"]
      #gets all data where the season and hour match that of the missing value
      season_hour_data = filter(data, Season == null_data[i, "Season"] & Hour == hour) 
      #gets average value
      hourly_average = mean(season_hour_data[,col_name], na.rm = TRUE)

      if(col_name == "RentCount")
      {
        hourly_average = floor(hourly_average)
      }
      #replaces the missing value with the average value
      data[rownames(null_data)[i],col_name] <- hourly_average
      filled_null_data[rownames(null_data)[i],col_name] <- hourly_average
    }
  }
}
```

Understanding the data
===================================== 

### Understanding The Data

Row {data-height=100}
-------------------------------------
### What can you say about the various variables?

The data given contains both numerical and categorical variables. The Categorical Variables include Date (Nominal), Hour (Ordinal), Season (Nominal), Holiday (Nominal), and Functional Day (Nominal). The Numerical Variables include Rented Bike Count (Discrete), Temperature (Continuous), Humidity (Continuous), Windspeed (Continuous), Visibility (Continuous), Solar Radiation (Continuous), Rainfall (Continuous), and Snowfall (Continuous).

Row {data-height=150}
-------------------------------------
### What statistical analysis would you apply over the data to understand it better?
To understand the data given, various statistical analysis can be performed. Descriptive Statistics can be applied to summarize the central tendency, dispersion, and shape of the distribution of the variables. This will help in getting an overview of the data and identifying any outliers. Correlation Analysis can be used to examine the relationship between different variables and determine if any of them are related to the rental bike count. This can help in identifying variables that are influential in affecting the rental bike count. Box Plots and Histograms can be used to visualize the distribution of the rental bike count and other variables. This can help in identifying the distribution of the data and any patterns that may be present.

Row {data-height=100}
-------------------------------------
### What can you say about the distribution of the Rented Bike Count variable?
From the histogram and based on the data given in the excel file, the rental bike count distribution appears to be skewed to the right, therefore, being a positive one, with more rented bikes in the lower intervals and fewer rented bikes in the higher intervals. The shape of the distribution is not symmetrical, indicating that the rental bike count may not follow a normal distribution.

Row {data.height=500}
-------------------------------------
### Histogram

```{r fig.width=18, fig.height=6}
histogram_data <- hist(data$RentCount, 
   col="#74b6f3", ylim=c(0,1400), xlim=c(0,3630), breaks=seq(0,3630,110), xaxp=c(0,3630,33), las = 2,
  main="Histogram of Rent Count", xlab="Ranges", ylab="Frequency")
```

Importing the Data {data-navmenu="Cleaning the data"}
===================================== 

### Importing and Cleaning the data

Row {data-height=100}
-------------------------------------
### Importing the Data
```{r echo = TRUE, eval=FALSE}
data <- read.csv("C:/Users/Malcolm/Downloads/SeoulBikeData.csv", header=FALSE)
```

Row
-------------------------------------
### Cleaning and Processing the data
``` {r echo = TRUE, eval=FALSE}
# remove 1st row
data <- data[-c(1), ]
# rename columns
colnames(data) <- c("date","RentCount", "Hour", "Temp", "Hum", "Wind", 'Vis', "Dew","SolRad","Rain", "Snow","Season", "Holiday","Function")
data <- type.convert(data, as.is= TRUE, na.strings = "NA")

data$Function[data$Function == 'No'] <- FALSE
data$Function[data$Function == 'Yes'] <- TRUE

data$Holiday[data$Holiday == 'No Holiday'] <- FALSE
data$Holiday[data$Holiday == 'Holiday'] <- TRUE

data$date <- dmy(data$date)
#settings the correct type
data$RentCount <- as.integer(as.character(data$RentCount))
data$Hour <- as.numeric(as.character(data$Hour))
data$Temp <- as.numeric(as.character(data$Temp))
data$Hum <- as.numeric(as.character(data$Hum))
data$Vis <- as.numeric(as.character(data$Vis))
data$Dew <- as.numeric(as.character(data$Dew))
data$Wind <- as.numeric(as.character(data$Wind))
data$SolRad <- as.numeric(as.character(data$SolRad))
data$Rain <- as.numeric(as.character(data$Rain))
data$Snow <- as.numeric(as.character(data$Snow))
```

Data Imputation {data-navmenu="Cleaning the data"}
===================================== 

### Imputing the data

Row
-------------------------------------

### The data provided contained null values as can be seen below:

```{r}
kable(null_data, caption = "Null values")
```

As a data imputation strategy we opted to use the '(Rounded) Mean or Moving Average or Median Value' strategy. This implementation was done by getting the average value of the missing data of the same Season and Hour and replacing the missing value with it. After the data imputation was complete, the null values were replaced:

```{r}
kable(filled_null_data, caption = "Imputated Null values")
```

The code for the imputation is as follows:
```{r echo = TRUE, eval = FALSE}
for(i in 1:nrow(null_data)) 
{
  for(j in 1:ncol(null_data))
  {
    # if there is a null in the column we replace it with the seasons hourly average
    if(is.na(null_data[i,j]))
    {
      # gets column name of missing value
      col_name = (colnames(null_data)[j])
      #gets the hour where the missing value is
      hour = null_data[i, "Hour"]
      #gets all data where the season and hour match that of the missing value
      season_hour_data = filter(data, Season == null_data[i, "Season"] & Hour == hour) 
      #gets average value
      hourly_average = mean(season_hour_data[,col_name], na.rm = TRUE)

      if(col_name == "RentCount")
      {
        hourly_average = floor(hourly_average)
      }
      #replaces the missing value with the average value
      data[rownames(null_data)[i],col_name] <- hourly_average
    }
  }
}
```


Deviation {data-navmenu="Data Analysis"}
===================================== 

Row {data-height=500}
-------------------------------------

### Graph

```{r fig.width=10}
# Task3

# candle 3

# group data by month
monthly <- data %>%  group_by(month = lubridate::floor_date(date, "month")) %>%
  summarize(average = mean(RentCount), sd = sd(RentCount))

# convert from date to month name
monthly$month <- months(as.Date(monthly$month)) 

# order by month
monthly <- monthly %>% mutate(month = factor(month, levels=month.name)) %>% arrange(month)

# plot average and sd per month
p <- ggplot(monthly, aes(x=month, y=average)) + ylab("Rent Count") + xlab("Month") + ggtitle("Average and Standard Deviation of Rented Bike Count for each month") +
  geom_bar(stat="identity",fill="#74b6f3") +
  geom_errorbar(aes(ymin=average-sd, ymax=average+sd), width=.3) +
  geom_point(size=2)

p
```

Row
-------------------------------------

### Description
The chart represents the average rented bike count for each month in a year, with the standard deviation indicated. January has an average rented bike count of slightly less than 250. February has a slightly higher average compared to January, but still less than 250. March sees a sharp increase in average rented bike count, reaching 500. April and May also witness a steady increase in average rented bike count, reaching 750 and a bit less than 1000 respectively. June has the highest average rented bike count of the year, reaching 1250. The average rented bike count drops to 1000 in July and then decreases to 875 in August and October. September has an average rented bike count slightly less than 1000. November and December have an average rented bike count of 625 and 250 respectively, which is lower compared to the earlier months.

Seasonal Statistics {data-navmenu="Data Analysis"}
===================================== 

Row {data-height=500}
-------------------------------------
### Graph

```{r fig.width=10}

# gets list of the different Seasons in the dataset
seasons <- unique(data$Season)

# creates empty dataframe
seasonal_hourly = data.frame()

# loops through the seasons
for(i in seasons)
{
  
  # filters the data to show only of season i
  x <- filter(data, Season == i) %>% 
    # groups the data by the hour
    group_by(Hour) %>% 
    # gets the mean RentCount for each hour
    summarise(average = mean(RentCount))
  
  #transposition of x 
  y <- transpose(x)
  # adds the season that corresponds to the data
  y <- cbind(y, i)
  # names the columns
  colnames(y) <- c(0:23, "Season")
  # removes the first row which held the column names
  y <- y[-c(1), ]
  # adds the season row to the seasonal hourly
  seasonal_hourly = rbind(seasonal_hourly, y)

}
#moves season column to front (not necessary but easier to understand when viewing)
seasonal_hourly <- seasonal_hourly %>% relocate(Season)
# melts the seasonal_hourly dataframe so as to be able to plot
melted <- melt(seasonal_hourly, id.vars="Season")
# plots a line for each Season
p <- melted %>% ggplot(aes(x=variable,y=value, color = Season, group=Season)) + xlab("Hour") + ylab("Average Rent Count") + ggtitle("Rented Bike Count by Hour of the day across season") + 
  geom_line(linewidth=1.2)

p
```

Row
-------------------------------------
### Description

This line graph represents the Rented Bike Count by the hour of the day across the season. From the graph, we can conclude that the most rented bike count was during the Summer, while the least was in Winter. All the lines have the same peaks which indicates that the busiest hours throughout all of the 4 seasons are 8 am and 6 pm.

Daily Statistics {data-navmenu="Data Analysis"}
===================================== 

Row {data-height=500}
-------------------------------------
### Graph

```{r fig.width=10}
# selecting only required columns
weekday_data <- data %>% select(date, RentCount, Hour)
# convert date to day of week
weekday_data$date <- weekdays(weekday_data$date)
# remove saturday and sunday
weekday_data <- filter(weekday_data, date != "Saturday" & date != "Sunday")

# get the different days of week in the data
days_of_week <- unique(weekday_data$date)

# create a new empty dataframe
grouped_weekday_data = data.frame()

# loops through the different days
for(i in days_of_week)
{
  # filters data to show only of the day i
  x <- filter(weekday_data, date == i) %>%
    # groups data by the hour
    group_by(Hour) %>% 
    # gets mean for each hour
    summarise(average=mean(RentCount))
  
  # transposition of x
  y <-transpose(x)
  # adds the day that corresponds to the date
  y <- cbind(y,i)
  # renames columns
  colnames(y) <- c(0:23, "Day")
  # removes the first row which held the column names
  y <-y[-c(1),]
  # adds the day row to the grouped weekday data
  grouped_weekday_data = rbind(grouped_weekday_data, y)
}
#moves day column to front (not necessary but easier to understand when viewing)
grouped_weekday_data <- grouped_weekday_data %>% relocate(Day)
# melts the grouped_weekday_data so as to be able to plot
melted_grouped_weekday_data <- melt(grouped_weekday_data, id.vars="Day")

# plots line for each day of week
p <- melted_grouped_weekday_data %>% ggplot(aes(x=variable, y=value, color=Day, group=Day)) +
  xlab("Hour") + ylab("Average Rent Count") + ggtitle("Rented Bike Count by Hour of the day across weekdays") + 
  geom_line(linewidth=1.2)

p
```

Row
-------------------------------------

### Description
This line graph represents the Rented Bike Count by Hour of the day across weekdays. From the graph, we can conclude that the most rented bike count was on Friday, while the least was on Tuesday and Thursday. All of the lines have more or less the same shape with 2 peaks which indicate the busiest hours of the day. These hours, just like the graph of Rented Bike Count by the hour of the day across seasons are at 8 am and 6 pm.

Summer and Winter Comparisons {data-navmenu="Data Analysis"}
===================================== 


Row {data-height=500}
-------------------------------------

### Graph 1
```{r}
#filters and only keeps required columns
summer_winter_data <- data %>% filter(Season == "Summer" | Season =="Winter") %>% select(date, Hour, RentCount, Season)

# gets hourly averages
summer_winter_hourly_average <- summer_winter_data %>% group_by(Season) %>% summarise(average=mean(RentCount))

# gets totals for day
summer_winter_daily_average <- summer_winter_data %>% group_by(date) %>% summarise(Season=Season, total=sum(RentCount)) %>%
# gets daily averages
  group_by(Season) %>% summarise(average=mean(total))

# plots daily averages
p1 <- ggplot(summer_winter_daily_average, aes(x=Season, y=average, fill=Season)) + 
  ylab("Average Rent Count") + xlab("Season") + ggtitle("Daily Averages of Rented Bike Count for Summer and Winter") +
  geom_bar(stat = "identity")

p1
```

### Graph 2
```{r}
# plots hourly averages
p2 <- ggplot(summer_winter_hourly_average, aes(x=Season, y=average, fill=Season)) + 
  ylab("Average Rent Count") + xlab("Season") + ggtitle("Hourly Averages of Rented Bike Count for Summer and Winter") +
  geom_bar(stat = "identity")
p2
```

Row
-------------------------------------

### Description
Graph 1 which is a bar chart represents the daily averages of Rented Bike Count for Summer and Winter while graph 2 is a bar chart that represents the Hourly Averages of Rented Bike Count for Summer and Winter. From graph 1 we can conclude that during the Summer the daily average rented bike count is much higher than during that of Winter with summer having a value of about 25000 while winter having a value of about 5000. Similarly, from graph 2 we can conclude that the Hourly average of rented bike count is much higher during the Summer than during the Winter with summer having a value of about 1100 and winter having a value of about 230.